Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-10-28 1.Paper/Paper.Rmd
fix some typos

# Introduction
Over the past few years, data sharing efforts have substantially increased and researchers increasingly often have access to large combined datasets derived from electronic health records or from individual participant data (IPD). For example, the clinical practice research datalink (CRPD) [@herrett2015] is an electronic health record (EHR) dataset in the UK, and has been used in a variety of medical research, such as the evaluation of health policy and drug efficacy. A recent example of an IPD-MA is the emerging risk factor collaboration [@theemergingriskfactorscollaboration2007], where data were combined from approximately 1.1 million individuals across 104 studies to investigate associations of cardiovascular diseases with several predictors. Individuals in these large datasets tend to be clustered in centres, countries or studies, where they have been subject to similar healthcare processes, and are therefore more alike than individuals from another cluster. Sometimes, clusters may also differ in participant eligibility criteria, follow-up length, predictor and outcome definitions, or in the quality of applied measurement methods. Hence, correlation is likely to be present between observations from the same cluster, which can lead to differences or ‘heterogeneity’ between clusters regarding baseline patient characteristics and subsequent outcomes.

@@

Therefore, the aim of this paper is to develop a multilevel imputation method for incomplete continuous and binary variables that are MNAR. In section \ref{heckman} we provide an introduction to the Heckman model and its estimation, and we extended to a hierarchical setting. Then in the section \ref{methods} we define the main steps of the proposed imputation method. Posteriorly, in the section \ref{simulation} we provide the settings and results of a Montecarlo simulation study used to evaluate the performance of the proposed impuatation method. In section \ref{example} we illustrate the method using the survey information collected in different sub-districts in Uganda to estimate the prevalence of malaria in children. 
Finally we provide a discussion  in section \ref{discussion} about  the results, limitations and propose future extension of our method.

@@

#### Fit a meta-analysis model
In this step, we pool the parameters $\widehat{\theta_i}$ with a random-effects meta-analysis model using only the groups with observable information, i.e., those that are not systematically missing and have sufficient information to estimate the heckman model.  In particular, we pooled the $p$ coefficients of the $\beta^O$ outcome equation and estimated a multivariate random effects meta-analysis model with them, similarly we combined all $q$ coefficient parameters of the $\beta^S$ selection equation. We also performed a univariate random effects meta_analysis on $\sigma'$ the log-transformed parameter of $\sigma$ and another on $\rho'$ the fisher-transformed parameter of $\rho$.

@@

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation $X_i^O=[1,X_{1i},X_{2i}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction, $X_i^S=[1,X_{1i},X_{2i},X_{3i}]$. Then in case of a missing continuous variable, we calculate the latent variables  $y_{i}^*$ and $r^*_{i}$ as follows:
\begin{align*}
 y_{i}^*&= \beta_i^OX_i^O+\epsilon^O_{i} \\
 r_{i}^*&= \beta_i^SX_i^O+\epsilon^S_{i}
\end{align*}
Here we assumed that all coefficient parameters varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{hi}^O&= \beta_h^O + b_{hi}^O\\
 \beta_{hi}^S&= \beta_h^S + b_{hi}^S
\end{align*}
We used these fixed coefficients $\beta_h^O=\{0.3,1,1\}$  and $\beta_h^S=\{-0.8,1.3,-0.7,1.2\}$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we made that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in data set.
We assumed that random effects were independent within equations ($b_{h0}^O\indep b_{h1}^O \indep b_{h2}^O$ and $b_{h0}^S \indep b_{h1}^S \indep b_{h2}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b^O_{h} \\
b^S_{h}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\sigma_{bh}^2\begin{pmatrix}
1 & \rho*0.4 \\
\rho*0.4 &1 
\end{pmatrix}\right)
\]
with the parameters $\sigma_{b0}^2=\sigma_{b1}^2=\sigma_{b2}^2=0.4$.  We considered that the correlation parameter of the random effects between equations is a 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_3\sim N(0,0.2)$ assuming that the intracluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter adopts different values depending on the simulated missing mechanism (See below additional scenarios) .

@@

#### Estimand
The estimads were the parameter coefficients of the outcome equation $\beta^O={\beta^O_0,\beta^O_1,\beta^O_2}$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report in the estimated variance of the random effects and residual errors $\sigma_{b0}^2$,$\sigma_{b1}^2$,$\sigma_{b2}^2$,$\sigma_{e}^2$.

@@

Figure \ref{fig:rhocont} shows the results of simulations where the missing variable was continuous.In the MAR scenario, i.e. when $\rho=0$, all imputation methods provide similar unbiased estimates of the coefficient parameters, but as rho increases, i.e. the mechanism becomes MNAR, the estimates for the complete case analysis and the MAR-IPD imputation method turns more biased.
Overall, both Heckman-based imputation (1l. and 2l.) models provide less biased estimates of the coefficient parameters in MNAR scenarios, but the random effects estimates on the 1l.heckman are far away from the true values as no cluster information was considered at all.
 Regarding the coverage, the Heckman-IPD imputation method provides the best coverage values across all the $\rho$ scenarios with a coverage level close to the nominal 95\% interval. Even 2l.Heckman is not properly a randomization-valid (Rubin 1987b) method across all the $\rho$ values it leads to better results in terms of bias and coverage compared to the evalueated methods.

The precision of the Heckman IPD separated seems it is better than the Heckman IPD full version, specially in the MAR scenarios, when we checked simulation results it seems that the high variability on the full version is related to simulations in which singularity fit that it is an indication that the model is overfitted.

The Heckman-IPD method, by incorporating individual and marginal level information, allows a bias-variance trade off, thus, as it was observed in Table \ref{tab:rhotable} the Heckman IPD separated resulted in better estimates in terms of RMSE than the other methods evaluated.
In particular, the method provides an advantage over the Heckman-single method in systematic missingness scenarios, since the latter does not allow to specify any cluster information implicitly, i.e. by adding a group variable in the imputation model. This can be seen in the estimates of the random effects parameters of the Heckman-single model, which are more biassed than those of the other methods in which cluster information was included.
Moreover, the method could be affected by the ratio of the pooled data (See Annex Figure\ref{fig:continuous3}). To demonstrate this, we increased the variance of the random effects in the simulated data, reducing the intra-cluster correlation (ICC) from 0.33 to 0.23. This resulted in less biassed estimates for both coefficient and random effects parameters, along with narrower confidence intervals.

@@

To illustrate our method, following the article by @rugnao2019, we estimated the prevalence of parasitemia by subregion and by age after approximately 3 years of LLIN campaigns started.  We estimated parasitemia prevalence using 3 approaches: MCAR, MAR and MNAR.

@@

Under the MNAR assumption, we used the proposed 2l.Heckman method to impute missing test values.  The selection and outcome equation included the same predictor variables used under the MAR approach. 
In addition, we included a holiday indicator variable as ERV, which was calculated according to school vacation calendars and public holidays in Uganda in 2017. We examined the association of this ERV with the outcome variable (y) and with the selection indicator (ry), conditionated on the remaining imputation predictors. The model results in Table \ref{tab:exclusion} indicate that the holiday indicator could be a plausible ERV variable, as it was significantly associated with ry, but not with y.

@@

print(xt, comment = FALSE, include.rownames=FALSE,add.to.row = comment, 
      hline.after = c(-1,0))
```
According to our imputation approach, non-participants were estimated to have a higher prevalence of malaria than participants in more than half of the districts analyzed. 
As can be seen in Figure \ref{fig:region} in terms of subregion level prevalence the estimates of the approaches do not differ significantly between methods, however except for the East-Central region, prevalence estimates under the Heckman approach are higher than those estimated using MAR or CC aproaches. 
```{r region,fig.pos="!H",fig.cap="\\label{fig:region} Estimates of malaria prevalence by Uganda subregion",echo=FALSE,fig.keep="last"}
plot_dist
```

@@

One of the major limitations of our method is that it needs a valid restriction variable, which in some contexts is difficult to establish at the individual study level and can be even more challenging if one tries to find a valid exclusion variable across clusters.
Also the method is sensitive to the value of the correlation between the selection equation and outcome (rho), and in general it is observed that it can lead to biassed results on fixed global parameters i.e. without variation across clusters.Similarly, the method can be sensitive to both the sample size and the number of studies included in the database. On the one hand a small sample size at the individual study level can affect not only the precision of estimates but also the convergence of the method since it requires a minimum sample size to estimate all the parameters of the Heckman model which can be at least twice the number of parameters required in a MAR prediction model. On the other hand, a high number of studies that may improve the precision of the estimators may also make the estimation of the marginal parameters more difficult and also considerably increase the processing time of our method.

>>>>

# Introduction
Over the past few years, data sharing efforts have substantially increased and researchers increasingly often have access to large combined datasets derived from electronic health records or from individual participant data (IPD). For example, the clinical practice research datalink (CRPD) [@herrett2015] is an electronic health record (EHR) dataset in the UK, and has been used in a variety of medical research, such as the evaluation of health policy and drug efficacy. A recent example of an IPD-MA is the emerging risk factor collaboration [@theemergingriskfactorscollaboration2007], where data were combined from approximately 1.1 million individuals across 104 studies to investigate associations of cardiovascular diseases with several predictors. Individuals in these large datasets tend to be clustered in centers, countries or studies, where they have been subject to similar healthcare processes, and are therefore more alike than individuals from another cluster. Sometimes, clusters may also differ in participant eligibility criteria, follow-up length, predictor and outcome definitions, or in the quality of applied measurement methods. Hence, correlation is likely to be present between observations from the same cluster, which can lead to differences or ‘heterogeneity’ between clusters regarding baseline patient characteristics and subsequent outcomes.

@@

Therefore, the aim of this paper is to develop a multilevel imputation method for incomplete continuous and binary variables that are MNAR. In section \ref{heckman} we provide an introduction to the Heckman model and its estimation, and we extended to a hierarchical setting. Then in the section \ref{methods} we define the main steps of the proposed imputation method. Posteriorly, in the section \ref{simulation} we provide the settings and results of a Monte Carlo simulation study used to evaluate the performance of the proposed imputation method. In section \ref{example} we illustrate the method using the survey information collected in different sub-districts in Uganda to estimate the prevalence of malaria in children. 
Finally we provide a discussion  in section \ref{discussion} about  the results, limitations and propose future extension of our method.

@@

#### Fit a meta-analysis model
In this step, we pool the parameters $\widehat{\theta_i}$ with a random-effects meta-analysis model using only the groups with observable information, i.e., those that are not systematically missing and have sufficient information to estimate the Heckman model.  In particular, we pooled the $p$ coefficients of the $\beta^O$ outcome equation and estimated a multivariate random effects meta-analysis model with them, similarly we combined all $q$ coefficient parameters of the $\beta^S$ selection equation. We also performed a univariate random effects meta_analysis on $\sigma'$ the log-transformed parameter of $\sigma$ and another on $\rho'$ the fisher-transformed parameter of $\rho$.

@@

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation $X_i^O=[1,X_{1i},X_{2i}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction, $X_i^S=[1,X_{1i},X_{2i},X_{3i}]$. Then in case of a missing continuous variable, we calculate the latent variables  $y_{i}^*$ and $r^*_{i}$ as follows:
\begin{align*}
 y_{i}^*&= \beta_i^OX_i^O+\epsilon^O_{i} \\
 r_{i}^*&= \beta_i^SX_i^O+\epsilon^S_{i}
\end{align*}
Here we assumed that all coefficient parameters varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{hi}^O&= \beta_h^O + b_{hi}^O\\
 \beta_{hi}^S&= \beta_h^S + b_{hi}^S
\end{align*}
We used these fixed coefficients $\beta_h^O=\{0.3,1,1\}$  and $\beta_h^S=\{-0.8,1.3,-0.7,1.2\}$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we made that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in data set.
We assumed that random effects were independent within equations ($b_{h0}^O\indep b_{h1}^O \indep b_{h2}^O$ and $b_{h0}^S \indep b_{h1}^S \indep b_{h2}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b^O_{h} \\
b^S_{h}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\sigma_{bh}^2\begin{pmatrix}
1 & \rho*0.4 \\
\rho*0.4 &1 
\end{pmatrix}\right)
\]
with the parameters $\sigma_{b0}^2=\sigma_{b1}^2=\sigma_{b2}^2=0.4$.  We considered that the correlation parameter of the random effects between equations is a 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_3\sim N(0,0.2)$ assuming that the intra-cluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter adopts different values depending on the simulated missing mechanism (See below additional scenarios) .

@@

#### Estimand
The estimands were the parameter coefficients of the outcome equation $\beta^O={\beta^O_0,\beta^O_1,\beta^O_2}$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report in the estimated variance of the random effects and residual errors $\sigma_{b0}^2$,$\sigma_{b1}^2$,$\sigma_{b2}^2$,$\sigma_{e}^2$.

@@

Figure \ref{fig:rhocont} shows the results of simulations where the missing variable was continuous.In the MAR scenario, i.e. when $\rho=0$, all imputation methods provide similar unbiased estimates of the coefficient parameters, but as rho increases, i.e. the mechanism becomes MNAR, the estimates for the complete case analysis and the MAR-IPD imputation method turns more biased.
Overall, both Heckman-based imputation (1l. and 2l.) models provide less biased estimates of the coefficient parameters in MNAR scenarios, but the random effects estimates on the 1l.Heckman are far away from the true values as no cluster information was considered at all.
 Regarding the coverage, the Heckman-IPD imputation method provides the best coverage values across all the $\rho$ scenarios with a coverage level close to the nominal 95\% interval. Even 2l.Heckman is not properly a randomization-valid (Rubin 1987b) method across all the $\rho$ values it leads to better results in terms of bias and coverage compared to the evaluated methods.

The precision of the Heckman IPD separated seems it is better than the Heckman IPD full version, specially in the MAR scenarios, when we checked simulation results it seems that the high variability on the full version is related to simulations in which singularity fit that it is an indication that the model is over-fitted.

The Heckman-IPD method, by incorporating individual and marginal level information, allows a bias-variance trade off, thus, as it was observed in Table \ref{tab:rhotable} the Heckman IPD separated resulted in better estimates in terms of RMSE than the other methods evaluated.
In particular, the method provides an advantage over the Heckman-single method in systematic missingness scenarios, since the latter does not allow to specify any cluster information implicitly, i.e. by adding a group variable in the imputation model. This can be seen in the estimates of the random effects parameters of the Heckman-single model, which are more biased than those of the other methods in which cluster information was included.
Moreover, the method could be affected by the ratio of the pooled data (See Annex Figure\ref{fig:continuous3}). To demonstrate this, we increased the variance of the random effects in the simulated data, reducing the intra-cluster correlation (ICC) from 0.33 to 0.23. This resulted in less biased estimates for both coefficient and random effects parameters, along with narrower confidence intervals.

@@

To illustrate our method, following the article by @rugnao2019, we estimated the prevalence of parasitemia by sub-region and by age after approximately 3 years of LLIN campaigns started.  We estimated parasitemia prevalence using 3 approaches: MCAR, MAR and MNAR.

@@

Under the MNAR assumption, we used the proposed 2l.Heckman method to impute missing test values.  The selection and outcome equation included the same predictor variables used under the MAR approach. 
In addition, we included a holiday indicator variable as ERV, which was calculated according to school vacation calendars and public holidays in Uganda in 2017. We examined the association of this ERV with the outcome variable (y) and with the selection indicator (ry), conditioned on the remaining imputation predictors. The model results in Table \ref{tab:exclusion} indicate that the holiday indicator could be a plausible ERV variable, as it was significantly associated with ry, but not with y.

@@

print(xt, comment = FALSE, include.rownames=FALSE,add.to.row = comment, 
      hline.after = c(-1,0))
```
According to our imputation approach, non-participants were estimated to have a higher prevalence of malaria than participants in more than half of the districts analyzed. 
As can be seen in Figure \ref{fig:region} in terms of sub-region level prevalence the estimates of the approaches do not differ significantly between methods, however except for the East-Central region, prevalence estimates under the Heckman approach are higher than those estimated using MAR or CC approaches. 
```{r region,fig.pos="!H",fig.cap="\\label{fig:region} Estimates of malaria prevalence by Uganda subregion",echo=FALSE,fig.keep="last"}
plot_dist
```

@@

One of the major limitations of our method is that it needs a valid restriction variable, which in some contexts is difficult to establish at the individual study level and can be even more challenging if one tries to find a valid exclusion variable across clusters.
Also the method is sensitive to the value of the correlation between the selection equation and outcome (rho), and in general it is observed that it can lead to biased results on fixed global parameters i.e. without variation across clusters.Similarly, the method can be sensitive to both the sample size and the number of studies included in the database. On the one hand a small sample size at the individual study level can affect not only the precision of estimates but also the convergence of the method since it requires a minimum sample size to estimate all the parameters of the Heckman model which can be at least twice the number of parameters required in a MAR prediction model. On the other hand, a high number of studies that may improve the precision of the estimators may also make the estimation of the marginal parameters more difficult and also considerably increase the processing time of our method.
